---
title: "Clebson Cardoso Alves de Sá"
author: '2016751813'
output:
  html_notebook: default
  html_document: default
  pdf_document: default
---

## 1. Geometric Mean
The function bellow computes the geometric mean of a vector $x = \{x_1, x_2, \dots, x_n\}$. If any of the values within the vector are negative, the function gives a warning message and computes the geometric mean. To compute the geometric mean, the following equations is used:

$$
f(x) = \left( \prod_{i}^{n}x_{i} \right)^{\frac{1}{n}}
$$
The code is quite simple, firstly we evaluate if any value of the vector is negative. If that is the case a warning message is returned informing the problem. Secondly the built-in `prod` function is used to compute the product of all values. Finally, the obtained result is raised to the power of $\frac{1}{n}$. The code can be viewed bellow.

```{r}
GeometricMean <- function(values){
  if(sum(values < 0) > 0)
    warning("Negative values found.\n")
  return(prod(values)**(1/length(values)))
}
```

Let's make a few test cases for the `GeometricMean` function. We are going to use as example three test cases obtained from the [MathIsFun](https://www.mathsisfun.com/numbers/geometric-mean.html) website. As can be verified, all outputs are correctly computed.
```{r}
GeometricMean(c(2, 18))
GeometricMean(c(10, 51.2, 8))
GeometricMean(c(1, 3, 9, 27, 81))
```

***

## 2. Building a New Dataset
Lets build a dataset using the random generation normal distribution with mean equal to 3 and default standard deviation 1. To do that, it's necessary to simple use the `rnorm` function and append the obtained results into a matrix. The code to that is as follows: 

```{r}
set.seed(123)  # set the random seed generator
data <- data.frame(matrix(rnorm(10000, mean=3), 
                          ncol=25, 
                          dimnames=list(NULL, paste("X", 1:25, sep="."))))
```

Now let`s take a look at how the newly created data frame looks like, to do that let's print the first 5 lines of the data.
```{r}
head(data, 5)
```

Now that we know how the data looks like, we can play around with the data solving the following problems:

### 2.1 Geometric Mean
Compute the geometric mean function in each column of the data frame data. Let's use the let's use the function `GeometricMean` created on topic 1. As we can observe, a few outputs are warning messages, since there are negative values in a few of the column's.

```{r}
for (col_idx in 1:dim(data)[2]){
  print(GeometricMean(data[,col_idx]))
}
```



### 2.2 Standard Deviation
Lets compute the standard deviation for each column in the dataset. R already has a built-in function called `sd` that computes the standard deviation. Let's use a vector to hold the standard deviation for each column, and only after computing the value for each column print the standard deviation for each column. The code is shown bellow:

```{r}
standard_deviation = c()  # vectr
for(i in 1:dim(data)[2]){
  standard_deviation = c(standard_deviation, sd(data[,i]))
}

print(standard_deviation)
```

As we can observe above, each value is the standard deviation for a given column in the range[1, 25]. Therefore the standard deviation for the first column is 0.9690, for the second is 0.9942 and the column 25 has a standard deviation of 0.9824.



### 2.3 Total
Now let's compute the total sum for each column. Just like the previous question, we are going to use R summation function which is called `sum` to compute the total for all columns of the data frame. The values for each column will be held in a vector and printed out after the total for all columns are computed.

```{r}
total_sum = c()
for (i in 1:dim(data)[2]){
  total_sum = c(total_sum, sum(data[,i]))
}
print(total_sum)
```

Just like for the standard deviation, the first value is the total for the first column, the second for the second column and so-on.

The sum for the lines are just alike, where the first value is the sum for the first line, the second value for the second line and so-on.

```{r}
total_sum = c()
for (i in 1:dim(data)[1]){
  total_sum = c(total_sum, sum(data[i,]))
}
print(total_sum)
```



### 2.4 Conditional Selection
Now let's select a subset of the original data, considering  the restriction that the values for the column $X.1 > 3$ and the values for the column $X.20 < 3$. There a few possibilities of doing this, the simpler way is to use the built-in R function `subset`, which return slices of the data that meet the giving condition. The code is shown bellow:

```{r}
sub <- subset(data, data$X.1 > 3 & data$X.20 < 3)
dim(sub)  # return the dimension of the data frame (lines, columns)
```

The second way is to use R data slicing, which is shown in the code bellow:

```{r}
sub2 <- data[data$X.1 > 3 & data$X.20 < 3,]
dim(sub2) # return the dimension of the data frame (lines, columns)
```

As it's possible to see, both approaches indicates that after meeting the condition, only 102 lines of the data are left.


### 2.5 Replacing Column Names
Let's take a look of how the column names looks like:
```{r}
colnames(data)
```

To modify the column names, it's necessary to replace the current variable names using the function `colnames` by passing the new variable names. The fragment of code to do this is quite simple, and can be seen bellow: 


```{r}
new_col_names = sprintf("%s%d", "Var", 1:25)
colnames(data) <- new_col_names
head(data, 5)
```

As it's possible to see in the head for the dataset, all column names were modified for the $Var1, Var2, \dots, Var25$.

***

## 3. Iris Data Set
Reading the data set Iris and verifying the first 5 lines.
```{r}
iris = read.csv("./iris.csv")
head(iris, 5)
```

### 3.1 Obtaining help
The documentation for datasets that are well known and the built-in functions in R can be checked using either `?` or the `help` function. This is done in the following way:

```{r}
?iris
help("iris")
```

### 3.2 Boxplot
It's possible to plot a box-and-whisker plot by calling the `boxplot` function in R. If a dataframe is given, it plots all variables in a single plot. The code is shown bellow:

```{r}
boxplot(iris)
```

### 3.3 BoxPlot in Groups
To plot multiple charts in the same frame we can split the window using the R function `par(mfrow=c(lines,cols))`. The aggregation of groups can be combined using the Tilde Operator `~` to separate the left and right hand sides of the dataframe.
This operation is done for each attribute of the iris dataset grouped by the attribute Species.

```{r}
par(mfrow=c(2,2))
boxplot(iris$Sepal.Length ~ iris$Species, main="Sepal.Length ~ Species")
boxplot(iris$Sepal.Width ~ iris$Species, main="Sepal.Width ~ Species")
boxplot(iris$Petal.Length ~ iris$Species, main="Petal.Length ~ Species")
boxplot(iris$Petal.Width ~ iris$Species, main="Petal.Width ~ Species")
```


### 3.4 Histogram
To compute the histogram, R provides the function `hist`. In the following chart it's shown the histogram for each column of the data set iris. To plot the histogram for the column Species, was necessary to cast it's values to numeric.

```{r}
par(mfrow=c(2,3))
hist(iris$Sepal.Length, xlab = "Sepal.Length")
hist(iris$Sepal.Width, xlab = "Sepal.Widht")
hist(iris$Petal.Length, xlab = "Petal.Length")
hist(iris$Petal.Width, xlab = "Petal.Width")

species = sapply(iris$Species, as.numeric)  # return the column species as numeric
hist(species, xlab = "Species")
```

***

## 4 Attitude Dataset
Let's read the data set attitude and check-out the 5 first lines of the data.

```{r}
attitude = read.csv('./attitude.csv')
head(attitude, n=5)
```




### 4.1 Discovering the dataset
To visualize the dataset information we can check the R documentation in the same way we did for the Iris collection.

```{r}
?attitude
help(attitude)
```


### 4.2 Plotting the Data
To plot a scatter plot of all attributes against all atributes we can call the `plot` function within R. In doing so, a $N \times N$ matrix of all atributes is ploted. The result of this plot can ve visualized bellow:

```{r fig.align='center'}
plot(attitude)
```

### 4.3 Computing the Mean
There are basically two ways of computing the mean of columns. One was already explained in exercises 2.2 (Standard Deviation) and 2.3 (Total Sum), which basically consists of looping through each column and then applying the function to be computed upon the rows. The code for this approach is shown bellow:

```{r}
mean_list = c()
for(i in 1:dim(attitude)[2])
  mean_list = append(mean_list, mean(attitude[,i]))
print(mean_list)
```

A tider and simpler way of doing this consist of using the `sapply` function within R. This approach perform the same operation, but much more faster. The code can be seen bellow:
```{r}
means_attitude <- sapply(attitude, mean)
print(means_attitude)
```

As can be seen, both operation are correctly computed with the same output for each column.


### 4.4 Transforming a variable
It's possible to easily transform the values of a column to discrete values with the command `cut`. This function basicaly divides the values in intervals considering the buckets given as slicing parameters. To update the dataframe, the only thing required is to make an assignment of the discrete values into the desired column of the dataframe. The code bellow shows how this procedure is done. The first 5 lines of the dataset is shown, and as we can observe, the column `attitude$complaints` is now categoric.



```{r}
attitude$complaints = cut(attitude$complaints,
                          c(-1, 60, 80, 100), 
                          labels=c("bad", "okay", "good"))
head(attitude, 5)
```


### 4.5 Rating ~ Complaints
Let's visualize the box-and-whisker plot by grouping the dataset by the categorical column `attitude$complaints`. As we can observe, $50\%$ of the data representing *bad* complaints are between $(\approx43, \approx65)$ with a median of $50$. For the neutral complaints *okay*, the $50\%$ quantile of the ratings are distributed between $60$ and $71$ with a median of $\approx67$. The ratings for complaints considered *good* are scattered with $50\%$ of the data between $(\approx71, \approx80)$ with a median of $\approx77$.

```{r fig.align='center'}
boxplot(attitude$rating ~ attitude$complaints, xlab="Complaint", ylab="Rating")
```

### 4.6 Bar Chart
Let's see the frequency of each complaint by using the bar chart. The first thing to do is to compute the amount of complaints for each category. This is done by using the logical operator of equivalence. Every outcome that is `TRUE` is considered to be one, while `FALSE` is considered to be zero. Therefore, to get the amount of complaints of each categorical value, we just need sum up the outcome of each comparison. After that, we just call the `barplot` function to plot the chart. This procedure is shown bellow.

```{r fig.align='center'}
good <- sum(attitude$complaints=='good')
okay <- sum(attitude$complaints=='okay')
bad <- sum(attitude$complaints=='bad')
barplot(c(good, okay, bad), 
        names.arg = c('good', 'okay', 'bad'), 
        ylab='Frequency', 
        xlab='Complaint')
```



### 4.7 Modyfying Variable Names
To modify the names of the discrete variable `attitude$complaints` it is necessary to replace the old values for the new categories. This can be done by calling the `factor` function. The code for this procedure is shown bellow. As it's possible to see for the first 5 lines, the collumn `attitude$complaints` contains the new categories.

```{r}
attitude$complaints = factor(attitude$complaints, labels = c("Ruim", "Normal", "Bom"))
head(attitude, 5)
```

Now that the new categories were applied, the only thing remaining is to redo the plots considering the new labels for categorical variable.


```{r fig.align='center'}
par(mfrow = c(1, 2))
boxplot(attitude$rating ~ attitude$complaints, xlab="Complaint", ylab="Rating")

good <- sum(attitude$complaints=='Bom')
okay <- sum(attitude$complaints=='Normal')
bad <- sum(attitude$complaints=='Ruim')
barplot(c(good, okay, bad), 
        names.arg = c('Bom', 'Normal', 'Ruim'), 
        ylab='Frequency', 
        xlab='Complaint')
```



